### 2018-03-13 paper/3-compute.tex
Stuff from Simon

\section{Build tasks, abstractly}\label{sec-compute}

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in the previous
section~\S\ref{sec-background} and design complex build systems from simple
primitives. Specifically, we present the \emph{compute} and \emph{build}
abstractions:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
\end{minted}
\vspace{1mm}

\noindent
\hs{Compute} represents an abstract collection of \emph{build tasks}, for
example, build rules in software build systems or formulas in spreadsheets. It
is completely isolated from the world of compilers, calculation chains, file
systems, caches, and all other complexities of real build systems. We explain
the compute type in~\S\ref{sec-compute-applicative}, discussing several
noteworthy special cases: functorial, applicative and monadic types of compute.

@@

We start by introducing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.
% We will use this vocabulary throughout the rest of the paper.

\emph{Keys} are used to distinguish \emph{values}. In software build systems,
keys are typically filenames, e.g.~\cmd{main.c}, whereas values are file
contents (a C program source code in this case). In spreadsheets, keys are cell
names, e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In Haskell code, we will use type variables \hs{k}
and \hs{v} to denote keys and values, respectively.

We use a cryptographic \emph{hash function} \hs{hash :: v -> Hash} for efficient
tracking and sharing of~build results, where \hs{Hash} is an abstract data type
equipped with the equality test.

A \emph{store} associates keys to values. It is convenient to assume that a
store is total function \store, i.e. it contains a value for every possible key.
We therefore also assume that the type of values is capable of encoding values
corresponding to non-existent keys (missing files or empty cells). This, in
particular, suggests that it is possible to \emph{depend on the absence of a
value}, which is a useful feature for build systems. In addition to usual
read and write operations, some stores also support a dedicated \emph{read hash}
operation, which is implemented more efficiently than by reading and hashing the
result. For example, GVFS (Git Virtual File System)~\cite{gvfs} stores file
hashes and downloads file contents only on demand -- an important feature for
cloud build systems.

@@

\subsection{Compute}\label{sec-compute-applicative}

We describe build tasks and their dependencies using functions of type
\hs{Compute}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Compute c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
\vspace{1mm}

\noindent
A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
cannot be computed, or \hs{Just}~\hs{value} obtained by \hs{fetch}ing the
dependencies of the \hs{key} and performing the required computation. A compute
must be polymorphic in \hs{f} but may request certain computational
power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}
by specifying the constraint~\hs{c}. Let us look at an example by recalling the
following spreadsheet from~\S\ref{sec-background-excel}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
B1 = A1 + A2
\end{minted}
\vspace{1mm}

\noindent
The only build task in this spreadsheet can be expressed as an \emph{applicative
compute}, operating on keys (cell names) of type \hs{String} and values of
type \hs{Integer}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
add :: Compute Applicative String Integer
add fetch key | key /= "B1" = Nothing
              | otherwise   = Just $ (+) <$> fetch "A1" <*> fetch "A2"
\end{minted}
\vspace{1mm}

\noindent
The cells \cmd{A1} and \cmd{A2} are inputs, which is indicated by returning
\hs{Nothing}. The cell \cmd{B1} is an output, whose value is computed by
fetching the values of \cmd{A1} and \cmd{A2} and applying the pure
function~\hs{(+)} using the methods of the \hs{Applicative} type
class~\cite{mcbride2008applicative}.

@@

\subsection{Build}\label{sec-general-build}

Equipped with the \hs{Compute} abstraction, we can now introduce \hs{Build} in
more detail. Recall the informal definition of a build system
from~\S\ref{sec-background}:

\vspace{2mm}
\emph{Definition (Build system).}\hspace{1.5mm}Given a description of tasks, a
\emph{build system} executes them in an order that respects their dependencies
to produce a specified output from supplied inputs.
\vspace{2mm}

\noindent
To make the definition more precise, we express it as the following function:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
\end{minted}
\vspace{1mm}

\noindent
The type closely matches the above definition: \hs{Compute c k v} is `a
description of tasks', a key~\hs{k} is~`a~specified output', and the store
\store holds `supplied inputs'. The additional input parameter \hs{Maybe i}
may provide auxiliary information from the previous build, as discussed
in~\S\ref{sec-background-summary}. The build system produces auxiliary
information \hs{i} for the next build as well as an updated store, where the
output \hs{k} must be up-to-date. We formally capture the correctness of the
resulting store in the following definition.

\definition[Correctness]{Let \hs{build :: Build c i k v} be a build system,
\hs{compute} describe an acyclic collection of build tasks, \hs{key} be an output
key, \hs{mi :: Maybe i} be auxiliary information, and \hs{store} be a store.
Then the \hs{result}ing store \hs{(_, result) = build compute k mi store}
is~\emph{correct} if there exists an \hs{ideal} store, such that

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{compute}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{compute} matches the value in the \hs{ideal} store:
    \[
    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
    that the output \hs{key} has the correct value, but do not impose any
    restrictions on intermediate keys, therefore permitting shallow cloud builds.
    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies. This means
    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
    store has constraints not only on the output, but also on the inputs.
\end{itemize}
}
\label{def-correct}

>>>>

\section{Build systems, abstractly}\label{sec-compute}

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in the previous
section~\S\ref{sec-background} and design complex build systems from simple
primitives. Specifically, we present the \emph{compute} and \emph{build}
abstractions, which are described in
Sections~\ref{sec-task} and \ref{sec-general-build} respectively.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}

@@

\begin{figure}
\begin{minted}{haskell}
-- Stores
data Store i k v   -- Abstract
getInfo  :: Store i k v -> i
putInfo  :: Store i k v -> i -> Store i k v
getValue :: Store i k v -> k -> v
putValue :: Hashable v => Store i k v -> k -> v -> Store i k v
getHash  :: Hashable v => Store i k v -> k -> Hash

-- Hashing
hash :: Hashable a => a -> Hash

-- Applicative
(<*>) :: Applicative f => f (a->b) -> f a -> f b  -- Left-associative
(<$>) :: Applicative f => (a->b) -> f a -> f b    -- Left-associative
pure  :: Applicative f => a -> f a

-- State monad
data State s
instance Applicative State
instance Monad       State
evalState :: State s a -> s -> s
getState  :: State s s
\end{minted}
\caption{Signatures of library functions}
\label{fig-lib-funs}
\end{figure}

We start by establishing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.
% We will use this vocabulary throughout the rest of the paper.

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up to date.  The store is a key/value mapping, togther with
\emph{persistent build state} that is maintained by the build system itself.

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.

@@

The types of the library functions we use in this paper is
given in Figure~\ref{fig-lib-funs}.
We use a cryptographic \emph{hash function} \hs{hash :: Hashable v =>
v -> Hash} for efficient tracking and sharing of~build results, where
\hs{Hash} is an abstract data type equipped with the equality test.

A \emph{store}, of type \hs{Store i k v}, associates keys to values.
It is convenient to assume that a
store is total function \store, i.e. it contains a value for every possible key.
We therefore also assume that the type of values is capable of encoding values
corresponding to non-existent keys (missing files or empty cells). This, in
particular, suggests that it is possible to \emph{depend on the absence of a
value}, which is a useful feature for build systems. In addition to usual
read and write operations, some stores also support a dedicated \emph{read hash}
operation, which is implemented more efficiently than by reading and hashing the
result. For example, GVFS (Git Virtual File System)~\cite{gvfs} stores file
hashes and downloads file contents only on demand -- an important feature for
cloud build systems.

The store also contains a single persistent value, of type \hs{i}, representing
the persistent state maintained by the build system itself -- its ``memory''.

\subsection{The Task abstraction}\label{sec-task}

Our first key idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type is best introduced through an example.
Consider this Excel spreadsheet (c.f. \S\ref{sec-background-excel}):
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
  A1: 10     B1: =A1 + A2
  A2: 20     B2: =B1 * 2
\end{minted}
\vspace{1mm}
Here cell \hs{A1} contains the value \hs{10} while cell \hs{B1} contains
the formula \hs{=A1+A2}.
We can represent the formulae of of this spreadsheet
with the following task description
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Compute Applicative String Integer
sprsh1 get "B1" = Just ((+) <$> get "A1" <*> "A2")
sprsh1 get "B2" = Just ((* 2) <$> get "B1")
sprsh1 _   _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the constant values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{get} function and a key. It pattern-matches on the key to see if it has
a task description (in the Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{get} to find the value of any keys on
which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures
are given in Figure~\ref{fig-lib-funs}.  We will return to the need for this
generality in
Section~\ref{sec-deps}\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}.

For now, we content ourselves with observing that a task description,
of type \hs{Task c k v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a call-back (\hs{get}) to get the values
of its dependencies.

\subsection{The Build abstraction}\label{sec-general-build}

Next comes our second key abstraction: a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date.  (What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\emph{SLPJ: is this right?}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dumbBuild :: Build Monad () k v
dumbBuild task k store = execState (go k) store
  where
    go :: k -> State s v
    go k = case (task go k) of
             Nothing  -> do { s <- getState; return (getValue s k) }
             Just act -> do { v <- act; updState (\s -> putStore s k v); return v }
\end{minted}
\vspace{1mm}
\emph{SLPJ: explain}

\subsection{Correctness of a build system} \label{sec-build-correctness}

We can now say what it means for a build system to be \emph{correct}, something
that is seldom stated formally:

\definition[Correctness]{Suppose \hs{build :: Build c i k v} is a build system,
\hs{task :: Task c k v} describes an acyclic collection of build tasks, \hs{key} is an output
key, \hs{store} be a store.
Then the \hs{result}ing store \hs{result = build compute key store}
is~\emph{correct} if there exists a store \hs{ideal}, such that

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{compute} matches the value in the \hs{ideal} store:
    \[
    \hs{execute}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.
    \hs{result}~\hs{key}~\hs{==}~\hs{ideal}~\hs{key}. In other words, we require
    that the output \hs{key} has the correct value, but do not impose any
    restrictions on intermediate keys, therefore permitting shallow cloud builds.
    \item The \hs{store}, \hs{ideal} and \hs{result} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies. This means
    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
    store has constraints not only on the output, but also on the inputs.
\end{itemize}
}
\label{def-correct}

\subsection{Computing dependencies}

@@
